---
title: "Cyclist_Data_Case_Study"
author: "Yash"
date: "24/12/2021"
output:
  html_document:
    df_print: paged
---

## Calling In Required packages

For performing various cleaning , analising and visualizing tasks we need to load some basic or advanced packages.
 
```{r}
 
library(tidyverse)
library(lubridate)
library(janitor)
library("dplyr")
rm(list=ls())  ##for clearing environment page

```

## Importing Files and Preparing Data

This section shows the process of importing data and preparing it for further cleaning and analysis.

```{r}
###
### Importing Data
###

df1<- read.csv("202004-divvy-tripdata.csv")
df2<- read.csv("202005-divvy-tripdata.csv")
df3<- read.csv("202006-divvy-tripdata.csv")
df4<- read.csv("202008-divvy-tripdata.csv")
df5<- read.csv("202009-divvy-tripdata.csv")
df6<- read.csv("202010-divvy-tripdata.csv")
df7<- read.csv("202011-divvy-tripdata.csv")
df8<- read.csv("202012-divvy-tripdata.csv")
df9<- read.csv("202101-divvy-tripdata.csv")
df10<- read.csv("202102-divvy-tripdata.csv")
df11<- read.csv("202103-divvy-tripdata.csv")
df12<- read.csv("202104-divvy-tripdata.csv")


ride_data <- rbind(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)


```

### Preparing Data

Now we will prepare data by creating some new required fields and changing some the one datatype into the required ones.

```{r}

ride_data$dates <- as.Date(ride_data$started_at)
ride_data$started_at <- ymd_hms(ride_data$started_at)
ride_data$ended_at <- ymd_hms(ride_data$ended_at)
```

```{r}
ride_data$start_hour <- hour(ride_data$started_at)
ride_data$end_hour <- hour(ride_data$ended_at)
ride_data$duration <- (ride_data$ended_at)-(ride_data$started_at)
ride_data$hours <- difftime(ride_data$ended_at,ride_data$started_at,units = c("hours"))
ride_data$minutes <- difftime(ride_data$ended_at,ride_data$started_at,units = c("mins"))
```

## Cleaning Data

Here we will clean our data by removing any unwanted or empty fields and deleting rows or columns with NA/null values.

```{r}
ride_data <- remove_empty(ride_data,which = c("rows","cols"))
ride_data$start_station_name[ride_data$start_station_name == ""]<- NA
ride_data <- ride_data %>% 
  drop_na()
```

```{r}
ride_data <-ride_data %>% 
  filter(duration > 0)

ride_data <- ride_data %>% select(dates,rideable_type,started_at,ended_at,start_station_name,member_casual,start_hour,end_hour,hours,minutes,duration)
```

## Change Log

This section highlights all the changes that have been made to process,clean and perapre data for analysis and draw some useful insights.

* initially we imported data of 12 months (04/2020 - 04/2021)

* then we created one single dataframe of those different 12 months data.

* further on we changed the datatype of "started_at" and "ended_at" columns from "char" to "ymd_hms" so it can be accessed to use as date and time formatting for creating further columns required by us.

* we created separate columns for dates , start_hour , end_hour , duration in secs,minutes,hours respectively.

* moving on we cleaned data by filtering where duration was less than zero , dropping rows and columns wth na values or empty ones.

* thus creating a final ride_data named dataframe for analysis.

## Analyising Data

Here we will analyse trends , make some observations by visualizing the processed,cleaned and prepared data.

```{r}
###
### Create a weekly dataframe relating start hours.
###
ride_data_2 <- ride_data %>% 
  group_by(Weekly = floor_date(dates,"week"),start_hour) %>% 
  summarise(total_mins = sum(minutes),
            Mean = mean(minutes),
            Median = median(minutes),
            Max = max(minutes),
            Min = min(minutes),
            Count = n())
```

```{r}
summary(ride_data_2$Count)
```

```{r}
xtabs(ride_data_2$Count~ride_data_2$start_hour)
```

Here is the plot showcasing weekly trend of number of rides throught our 12 month data.
```{r}
ride_data_2 %>% 
  ggplot()+geom_col(mapping=aes(x=Weekly,y=Count))+
  ggtitle("Count of Rides Per Week")
```

Plot showing number of rides related to hours to analysise most active hours
```{r}
ride_data_2 %>% 
  ggplot()+geom_col(mapping=aes(x=start_hour,y=Count,fill=Count))+
  ggtitle("Count Of Rides by hour")
```

Creating dataframe relating number of rides with customer type.
```{r}
ride_data_3 <- ride_data %>% 
  group_by(Weekly = floor_date(dates,"week"),member_casual) %>% 
  summarise(total_mins = sum(minutes),
            Mean = mean(minutes),
            Median = median(minutes),
            Max = max(minutes),
            Min = min(minutes),
            Count = n())
```
Plot showing trend between weeekly trends for number of rides by customer type.
```{r}
ride_data_3 %>% 
  ggplot()+geom_area(mapping=aes(x=Weekly,y=Count,fill=member_casual))+ggtitle("Count of Rides related to customer type")
```

```{r}
ride_data_3 %>% 
  ggplot()+geom_col(mapping=aes(x=Weekly,y=total_mins,fill=member_casual))+scale_y_continuous()+ggtitle("Total Amount of time spent by customer type") 
```